Vertical and oblique transmission with fluctuating transmission

### 2018-03-16 ms/fluctuating_oblique.tex
k-l result

There has been much less theory developed for fluctuating transmission.
Nevertheless, if we assume that social learning can be affected by weather, food availability, population density, or even political atmosphere, then it is reasonable that the mode and rate of transmission will fluctuate over time. 
For example, it may be the case that during cold seasons offspring are more likely to stay close to their parents and therefore learn vertically, whereas during warm seasons they are more likely to wander and interact with other adults and therefore learn obliquely.

Here, we analyze a model in which individuals learn from their parents with probability $\rho$ (vertical transmission), and from non-parental adults with probability $1-\rho$ (oblique transmission), where $\rho$ fluctuates over time~(\autoref{fig:transmission}).
We find that a phenotypic polymorphism cannot be maintained unless fluctuations in the mode of transmission are very small and only weakly correlated with fluctuations in selection.
We suggest that cultural evolution is affected by environmental changes that may affect both selection and transmission, and that further empirical and theoretical studies are required to disentangle these effects.

@@

% Model
\section*{Model}

Consider a very large population whose members are characterized by a single dichotomous cultural trait with phenotypes $A$ and $B$ and fitness values $w_A=1+s$ and $w_B=1$, respectively.
Phenotypes are transmitted vertically with probability $\rho$ or obliquely with probability $1-\rho$.
Given $x$ the frequency of phenotype $A$ at the current generation, the frequency of $A$ in the next generation is
\begin{equation} \label{eq:recurrence}
x' = \rho \frac{1+s}{\overline w} x + (1-\rho)x,
\end{equation}
where $\overline w = 1 + xs$ is the population mean fitness.

@@

If the environment changes every $k$ generations, so that for the first $k$ generations $A$ is favored and the transmission rate is $\rho_1$ and for the second $k$ generations $B$ is favored and the transmission rate is $\rho_2$.
The change in the frequency of $A$ when either $A$ or $B$ is favored is described by $F_A(x)$ and $F_B(x)$, respectively, where
\begin{equation}\begin{aligned} \label{eq:recurrence_periodic_fluc}
F_A(x) = \rho_1 \frac{1+s}{1+sx} x + (1-\rho_1)x, \\
F_B(x) = \rho_2 \frac{1}{1+s-sx'} x' + (1-\rho_2)x'.
\end{aligned}\end{equation}
Fixations of $A$ ($x^*=1$) and $B$ ($x^*=0$) are locally stable if, respectively,
\begin{equation}\begin{aligned}
\big(F'_A(1) F'_B(1)\big)^k < 1, \\
\big(F'_A(0) F'_B(0)\big)^k < 1. \\
\end{aligned}\end{equation}
and if these both of these conditions are not met then there exists a protected polymorphism.
Therefore, for a protected polymorphism we require
\begin{equation}
1 < F'_A(1) F'_B(1) = \big(1-\rho_1\frac{s}{1+s}\big)\big(1+\rho_2 s\big) 
= 1-\frac{1}{1+s}\big(\rho_2(1+s-\rho_1)-\rho_1\big),
\end{equation}
and
\begin{equation}
1 < F'_A(0) F'_B(0) = \big(1+\rho_1 s\big)\big(1-\rho_2 \frac{s}{1+s}\big) 
= 1+\frac{s}{1+s}\big(\rho_1(1+s) -\rho_2(1+\rho_1 s)\big),
\end{equation}
which can be summarized as a condition on $\rho_2$
\begin{equation} \label{eq:poly_condition_periodic_fluc}
\frac{\rho_1}{1+s(1-\rho_1)} < \rho_2 < \frac{(1+s)\rho_1}{1+s\rho_1},
\end{equation}
or as a condition on the difference between $\rho_1$ and $\rho_2$
\begin{equation}
-s\rho_2(1-\rho_1) < \rho_2 - \rho_1 < s\rho_1(1-\rho_2).
\end{equation}

@@

@@

% Results
%\section*{Results}

% Discussion
%\section*{Discussion}

>>>>

There has been much less theory developed for fluctuating transmission.
Nevertheless, if we assume that social learning can be affected by weather~\citep{Phithakkitnukoon2012}, food availability, population density, or even political atmosphere, then it is reasonable that the mode and rate of transmission will fluctuate over time. 
For example, it may be the case that during cold seasons offspring are more likely to stay close to their parents and therefore learn vertically, whereas during warm seasons they are more likely to wander and interact with other adults and therefore learn obliquely.

Here, we analyze a model in which individuals learn from their parents with probability $\rho$ (vertical transmission), and from non-parental adults with probability $1-\rho$ (oblique transmission), where $\rho$ fluctuates over time~(\autoref{fig:transmission}).
We find that $\ldots$ %TODO
We suggest that cultural evolution is affected by environmental changes that may affect both selection and transmission, and that further empirical and theoretical studies are required to disentangle these effects.

@@

% Model & Results
\section*{Model \& Results}

Consider a very large population whose members are characterized by a single dichotomous cultural trait with phenotypes $A$ and $B$ and fitness values $w_A=1+s$ and $w_B=1$, respectively.
Phenotypes are transmitted vertically with probability $\rho$ or obliquely with probability $1-\rho$~(\autoref{fig:transmission}).
Given $x$ the frequency of phenotype $A$ at the current generation, the frequency of $A$ in the next generation is
\begin{equation} \label{eq:recurrence}
x' = \rho \frac{1+s}{\overline w} x + (1-\rho)x,
\end{equation}
where $\overline w = 1 + xs$ is the population mean fitness.

@@

If the environment changes every $k$ generations, so that for the first $k$ generations $A$ is favored and the transmission rate is $\rho_1$ and for the second $k$ generations $B$ is favored and the transmission rate is $\rho_2$.
The change in the frequency of $A$ when either $A$ or $B$ is favored is described by $F_A(x)$ and $F_B(x)$, respectively, where
\begin{equation}\begin{aligned} \label{eq:recurrence_periodic_fluc}
F_A(x) = \rho_1 \frac{1+s}{1+sx} x + (1-\rho_1)x, \\
F_B(x) = \rho_2 \frac{1}{1+s-sx'} x' + (1-\rho_2)x'.
\end{aligned}\end{equation}
Fixations of $A$ ($x^*=1$) and $B$ ($x^*=0$) are locally stable if, respectively,
\begin{equation}\begin{aligned}
\big[F'_A(1) F'_B(1)\big]^k < 1, \\
\big[F'_A(0) F'_B(0)\big]^k < 1. \\
\end{aligned}\end{equation}
and if both of these conditions are not met then there exists a protected polymorphism.
Therefore, for a protected polymorphism we require
\begin{equation}
1 < F'_A(1) F'_B(1) = \Big(1-\rho_1\frac{s}{1+s}\Big)\Big(1+\rho_2 s\Big) 
= 1-\frac{1}{1+s}\big[\rho_2(1+s-\rho_1)-\rho_1\big],
\end{equation}
and
\begin{equation}
1 < F'_A(0) F'_B(0) = \Big(1+\rho_1 s\Big)\Big(1-\rho_2 \frac{s}{1+s}\Big) 
= 1+\frac{s}{1+s}\big[\rho_1(1+s) -\rho_2(1+\rho_1 s)\big],
\end{equation}
which can be summarized as a condition on $\rho_2$
\begin{equation} \label{eq:poly_condition_periodic_fluc}
\frac{\rho_1}{1+s(1-\rho_1)} < \rho_2 < \frac{(1+s)\rho_1}{1+s\rho_1},
\end{equation}
or as a condition on the difference between $\rho_1$ and $\rho_2$
\begin{equation}
-s\rho_2(1-\rho_1) < \rho_2 - \rho_1 < s\rho_1(1-\rho_2).
\end{equation}

@@

In the more general case, phenotype $A$ is favored for $k$ generations and phenotype $B$ for $l$ generations, and the transmission rate follows the same cycle such that $\rho=\rho_1$ when $A$ is favored and $\rho=\rho_2$ when $B$ is favored.
The requirements for a protected polymorphism is now
\begin{equation}\begin{aligned}
1 &< (F'_A(0))^k (F'_B(0))^l = a^k \big(b-\Delta \rho s/(1+s)\big)^l,\\
1 &< (F'_A(1))^k (F'_B(1))^l = b^k (a+\Delta \rho s)^l,
\end{aligned}\end{equation}
where $a=1+\rho_1 s>1$, $b=1-\rho_1\frac{s}{1+s}<1$, and $\Delta \rho = \rho_2 - \rho_1$.
This leads to a condition similar to eq.~20 in~\citet{Ram2018}, but more complex due to the addition of $\Delta \rho$:
\begin{equation} \label{eq:poly_condition_periodic_fluc_k_l}
\frac{-\log{b}}{\log{\big(a+\Delta \rho s\big)}} < 
\frac{l}{k} < 
\frac{\log{a}}{-\log{\big(b-\Delta \rho s/(1+s)\big)}}.
\end{equation}
Note that as $\Delta \rho$ increases, both the RHS and the LHS of \eqref{eq:poly_condition_periodic_fluc_k_l} decrease.
Therefore, increasing the vertical transmission rate specifically when $B$ is favored ($\Delta \rho>0$) decreases the environmental period ratio $l/k$ that permits protected polymorphism; reducing the vertical transmission rate when $B$ is favored ($\Delta \rho<0$) will have the opposite effect, increasing the ratio $l/k$ that permits a protected polymorphism.

@@

% Discussion
\section*{Discussion}

Models of cultural transmission mostly assume a fixed relative rate at which different modes of transmission -- vertical, horizontal, or oblique -- occur.
Here we explored a model in which the relative rate of vertical and oblique transmission fluctuates over time, either periodically or randomly. 
Comparing our results with previous results from a similar model with a fixed rate~\citep{Ram2018}, we find that transmission fluctuations constrain the maintenance of phenotypic polymorphism.
%Specifically, if fluctuations are not small then a single phenotype is likely to fix in the population~(\autoref{eq:poly_condition_periodic_fluc}, \autoref{fig:rho1_rho2_k=1}, \autoref{fig:rho1_rho2_stoch_p}), and

% Face book rural internet
% National service - Israel, Singapore
% Immigrtion - draught
% Opening schools, universities
% sending students abroad, bringing experts from abroad
% foreign labor
% refugees
% hunter-gatherers/chimps increase habitat in years with less food production
% religion - dark ages etc.
